Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-02-01 paper/main.tex
final pass through intro

\title{Text embedding models yield high resolution insights into conceptual
knowledge from short multiple choice quizzes}

@@

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple choice quiz questions interleaved between
watching two course videos from the Khan Academy platform. We applied our
framework to the videos' transcripts, and to text of the quiz questions, to
quantify the content of each moment of video and each quiz question. We used
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings show how a
small set of quiz questions may be used to obtain rich and meaningful high
resolution insights into what each learner knows, and how their knowledge
changes over time as they learn.

@@

Suppose that a teacher had access to a complete tangible ``map'' of everything
a student knew. Defining what such a map might even look like, let alone how it
might be constructed or filled in, is itself a non-trivial problem. But if a
teacher \textit{were} to gain access to such a map, how might it change their
ability to teach that student? Perhaps they might start by checking how well
the student knew the to-be-learned information already, or how much they knew
about related concepts. For some students, they could potentially optimize
their teaching efforts to maximize efficiency by focusing primarily on
not-yet-known content. For other students (or other content areas), it might be
more effective to optimize for direct connections between already known content
and new material. Observing how the student's knowledge changed over time, in
response to their teaching, could also help to guide the teacher towards the
most effective strategy for that individual student.

A common approach to assessing students' learning is to compile a set of quiz
questions, tally up the proportion of correctly answered questions, and provide
the student with feedback in the form of a letter grade. Although numerical or
letter grades give \textit{some} indication about whether the student has
mastered the to-be-learned material, any univariate measure of performance on a
complex task is at risk of conflating underlying factors, losing relevant
information, and so on. For example, consider the relative utility of the
imaginary map described above that characterizes in detail a student's
knowledge, versus a single annotation saying that the student answered 85\% of
their quiz questions correctly, or that they received a `B'. Here, we show that
the same quiz data required to compute proportion correct scores or letter
grades can be used to instead obtain much more detailed insights into what the
student knows at the time they took the quiz.

@@

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience~\citep[e.g.,][]{Kato40, Gall00, ScotEtal07, HallGree08, Macl05}
has profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
information about what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it. However, text
embedding models~\citep[e.g.,][]{LandDuma97, DeerEtal90, BleiEtal03,
BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20} also attempt to capture the
deeper meaning \textit{underlying} those atomic elements. These models consider
not only the co-occurrences of those elements within and across documents, but
also patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical properties
of the elements, and other high-level characteristics of how they are
used~\citep{Mann20, Mann21a}. According to these models, the deep conceptual
meaning of a document may be captured by a feature vector in a high-dimensional
representation space, where nearby vectors reflect conceptually related
documents. A model that succeeds at capturing an analog of ``understanding'' is
able to assign nearby feature vectors to two conceptually related documents,
\textit{even when the words contained in those documents have very little
overlap}.

@@

Here we develop a framework for modeling how knowledge is acquired during
learning. The central idea is to use text embedding models to define the
coordinate systems of two maps: (a) a \textit{knowledge map} that describes the
extent to which each concept is currently known and (b) a \textit{learning map}
that describes changes in knowledge over time. Each location on these maps
represents a single concept, and the maps' geometries are defined such that
related concepts are located nearby in space. We use this framework to analyze
and interpret behavioral data collected from an experiment that had
participants watch and answer multiple-choice questions about a series of
recorded course lectures.

Our primary research goal is to advance our understanding of what it means to
acquire deep, real-world conceptual knowledge. Traditional laboratory
approaches to studying learning and memory (e.g., list learning studies) often
draw little distinction between memorization and understanding. Instead, these
studies typically focus on whether information is effectively encoded or
retrieved, rather than whether the information is \textit{understood}.
Approaches to studying conceptual learning, such as category learning
experiments, can begin to investigate the distinction between memorization and
understanding, often by training participants to distinguish arbitrary or
random features in otherwise meaningless categorized stimuli. However the
objective of real-world training, or learning from life experiences more
generally, is often to develop new knowledge that may be applied in
\textit{useful} ways in the future. In this sense, the gap between modern
learning theories and modern pedagogical approaches and classroom learning
strategies is enormous: most of our theories about \textit{how} people learn
are inspired by experimental paradigms and models that have only peripheral
relevance to the kinds of learning that students and teachers actually
seek~\citep{Macl05, HallGree08}. To help bridge this gap, our study uses course
materials from real online courses to inform, fit, and test models of
real-world conceptual learning. We also provide a demonstration of how our
models can be used to construct ``maps'' of what students know, and how their
knowledge changes with training. In addition to helping to visualize knowledge
(and changes in knowledge), we hope that such maps might lead to real-world
tools for improving how we educate.

@@

    \caption{\textbf{Experimental paradigm.} Participants alternate between
    answering 13-question multiple-choice quizzes and watching two Khan Academy
    lectures. Each quiz contains a mix of 5 questions about lecture 1, 5
    questions about lecture 2, and 3 questions about general physics knowledge. The
    specific questions reflected on each quiz, and the orders of each quiz's
    questions, were randomized across participants.}

>>>>

\title{Text embedding models yield high-resolution insights into conceptual
knowledge from short multiple-choice quizzes}

@@

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple-choice quiz questions interleaved between
watching two course videos from the Khan Academy platform. We applied our
framework to the videos' transcripts and the text of the quiz questions to
quantify the content of each moment of video and each quiz question. We used
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings show how a
small set of quiz questions may be used to obtain rich and meaningful, high-resolution
insights into what each learner knows, and how their knowledge
changes over time as they learn.

@@

Suppose that a teacher had access to a complete, tangible ``map'' of everything
a student knew. Defining what such a map might even look like, let alone how it
might be constructed or filled in, is itself a non-trivial problem. But if a
teacher \textit{were} to gain access to such a map, how might it change their
ability to teach that student? Perhaps they might start by checking how well
the student knew the to-be-learned information already, or how much they knew
about related concepts. For some students, they could potentially optimize
their teaching efforts to maximize efficiency by focusing primarily on
not-yet-known content. For other students (or other content areas), it might be
more effective to optimize for direct connections between already known content
and new material. Observing how the student's knowledge changed over time, in
response to their teaching, could also help to guide the teacher towards the
most effective strategy for that individual student.

A common approach to assessing a student's knowledge is to present them with a set of quiz
questions, calculate the proportion they answer correctly, and provide
them with feedback in the form of a simple numeric or letter grade. While such a
grade can provide \textit{some} indication of whether the student has
mastered the to-be-learned material, any univariate measure of performance on a
complex task sacrifices certain relevant information, risks conflating underlying
factors, and so on. For example, consider the relative utility of the
imaginary map described above that characterizes a student's knowledge
in detail, versus a single annotation saying that the student answered 85\% of
their quiz questions correctly, or that they received a `B'. Here, we show that
the same quiz data required to compute proportion-correct scores or letter
grades can instead be used to obtain much more detailed insights into what the
student knows at the time they took the quiz.

@@

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience (e.g.,~\citealp{Kato40, Gall00, ScotEtal07, HallGree08, Macl05})
has profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
information about what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it. However, text
embedding models (e.g.,~\citealp{LandDuma97, DeerEtal90, BleiEtal03,
BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20}) also attempt to capture the
deeper meaning \textit{underlying} those atomic elements. These models consider
not only the co-occurrences of those elements within and across documents, but
also patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical properties
of the elements, and other high-level characteristics of how they are
used~\citep{Mann20, Mann21a}. According to these models, the deep conceptual
meaning of a document may be captured by a feature vector in a high-dimensional
representation space, where nearby vectors reflect conceptually related
documents. A model that succeeds at capturing an analog of ``understanding'' is
able to assign nearby feature vectors to two conceptually related documents,
\textit{even when the specific words contained in those documents have very little
overlap}.

@@

Here we develop a framework for modeling how conceptual knowledge is acquired during
learning. The central idea behind our framework is to use text embedding models to define the
coordinate systems of two maps: (a) a \textit{knowledge map} that describes the
extent to which each concept is currently known and (b) a \textit{learning map}
that describes changes in knowledge over time. Each location on these maps
represents a single concept, and the maps' geometries are defined such that
related concepts are located nearby in space. We use this framework to analyze
and interpret behavioral data collected from an experiment that had
participants answer sets multiple-choice questions about a series of
recorded course lectures.

Our primary research goal is to advance our understanding of what it means to
acquire deep, real-world conceptual knowledge. Traditional laboratory
approaches to studying learning and memory (e.g., list-learning studies) often
draw little distinction between memorization and understanding. Instead, these
studies typically focus on whether information is effectively encoded or
retrieved, rather than whether the information is \textit{understood}.
Approaches to studying conceptual learning, such as category learning
experiments, can begin to investigate the distinction between memorization and
understanding, often by training participants to distinguish arbitrary or
random features in otherwise meaningless categorized stimuli. However the
objective of real-world training, or learning from life experiences more
generally, is often to develop new knowledge that may be applied in
\textit{useful} ways in the future. In this sense, the gap between modern
learning theories and modern pedagogical approaches that inform classroom learning
strategies is enormous: most of our theories about \textit{how} people learn
are inspired by experimental paradigms and models that have only peripheral
relevance to the kinds of learning that students and teachers actually
seek~\citep{Macl05, HallGree08}. To help bridge this gap, our study uses course
materials from real online courses to inform, fit, and test models of
real-world conceptual learning. We also provide a demonstration of how our
models can be used to construct ``maps'' of what students know, and how their
knowledge changes with training. In addition to helping to visualize knowledge
(and changes in knowledge), we hope that such maps might lead to real-world
tools for improving how we educate.

@@

    \caption{\textbf{Experimental paradigm.} Participants alternate between
    completing three 13-question multiple-choice quizzes and watching two Khan Academy
    lectures. Each quiz contains a mix of 5 questions about lecture 1, 5
    questions about lecture 2, and 3 questions about general physics knowledge. The
    specific questions reflected on each quiz, and the orders of each quiz's
    questions, were randomized across participants.}
